---
title: "Handling Missing Data in Research: A Practical Guide"
subtitle  : "<small>Traditional Methods</small>"
author: "`r paste(params$author, collapse = ' & ')`"
date: "`r paste0(params$location, ', ', params$date)`"
params:
  author: ["Author Name"]
  date: "Placeholder Date"
  location: "Placeholder Location"
---
exclude: true

```{r child = "_setup.Rmd"}
```


---

#  Listwise (or pairwise) deletion

  - Drop cases with missing data
  - Default (in virtually every program)
  - .highlight[Requires MCAR]
  - Power loss (N decrease)

---

class: center, middle

.pull-right-60[

## Deletion `r emo::ji("raised_eyebrow")`

```{r, out.width = "450px", out.height = "425px", echo = F, message=FALSE, warning=FALSE, dpi = 300}
library(dplyr)
library(ggplot2)
library(smplot2)

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            45.0, 44.7, 48.5, 50.2, 42.8, 50.0, 54.0, 52.0, 58.0, 60.0)

# Create the main data frame
datac <- data.frame(ID, var1, var2)

# Calculate summary statistics
mean_var1c <- round(mean(var1, na.rm=T),2)
mean_var2c <- round(mean(var2, na.rm=T),2)
sd_var1c <- round(sd(var1, na.rm=T),2)
sd_var2c <- round(sd(var2, na.rm=T),2)

cor_test <- cor.test(var1, var2, use = "complete.obs")
correlc <- round(cor_test$estimate, 2)
if(cor_test$p.value < .001) {
  p_valuec <- "< .001"} else {
  p_valuec <- paste0("= ", round(cor_test$p.value, 3)) }


p1 <- ggplot(data = datac, mapping = aes(x = var1, y = var2)) +
  geom_point(size = 7, color = "#00A9E0", alpha = 0.6) + # Increase point size
  geom_smooth(method = "lm", se = FALSE, color = "#8DC63F", size = 3.5, alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
  sm_statCorr(color = "#8DC63F", text_size = 8) + 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 


# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)

# Create the main data frame
datam <- data.frame(ID, var1, var2)

# Calculate summary statistics
mean_var1m <- round(mean(var1, na.rm=T),2)
mean_var2m <- round(mean(var2, na.rm=T),2)
sd_var1m <- round(sd(var1, na.rm=T),2)
sd_var2m <- round(sd(var2, na.rm=T),2)

cor_test <- cor.test(var1, var2, use = "complete.obs")
correlm <- round(cor_test$estimate, 2)
if(cor_test$p.value < .001) {
  p_valuem <- "< .001"} else {
  p_valuem <- paste0("= ", round(cor_test$p.value, 3)) }

p2 <- ggplot(data = datam, mapping = aes(x = var1, y = var2)) +
  geom_point(size = 7, color = "#00A9E0", alpha = 0.6) + # Increase point size
  geom_smooth(method = "lm", se = FALSE, color = "#8DC63F", size = 3.5) +
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
  sm_statCorr(color = "#8DC63F", text_size = 8) + 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 

# Combine p1 and p2, adding p2 on top of p1

p3 <- ggplot(data = datam, mapping = aes(x = var1, y = var2)) +
  
  sm_statCorr(color = "white", text_size = 8) + 
  
  geom_point(data = datac, size = 7, color = "#00A9E0", alpha = 0.05) + # Increase point size
  #geom_smooth(data = datac, method = "lm", se = FALSE, color = "#8DC63F", size = 3.5, alpha = 0.2) +
  stat_smooth(data = datac, geom='line', method = "lm", alpha=0.08, se=FALSE, color = "#8DC63F", size = 3.5) +
  
  geom_point(data = datam, size = 7, color = "#00A9E0", alpha = 0.8) +
  #geom_smooth(data = datam, method = "lm", se = FALSE, color = "#8DC63F", size = 3.5) +
  stat_smooth(data = datam, geom='line', method = "lm", alpha=1, se=FALSE, color = "#8DC63F", size = 3.5) +
  
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 

p3

```
<small><small>.highlight[Complete Data Reference: *R* = `r correlc`, *p* `r p_valuec`]</small></small>

]

.pull-left-10[
```{r}

# for table
datam$var1 <- sprintf("%.2f", datam$var1)
datam$var2 <- ifelse(is.na(datam$var2), "", sprintf("%.2f", datam$var2))

# Create the table with kable and apply styling
knitr::kable(
  datam,
  format = "html",
  align = "ll",
  escape = FALSE
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    font_size = 13.5, full_width = TRUE
  )
```

]

---

#  Mean Imputation

  - Replace missing values with the average
  - Super simple and seems intuitive 
  - Can annihilate variability  
  - Likely the worst option

---

class: center, middle

.pull-right-60[

## Mean Imputation `r emo::ji("flushed")` 

```{r, out.width = "450px", out.height = "425px", echo = F, message=FALSE, warning=FALSE, dpi = 300}
library(dplyr)
library(ggplot2)
library(smplot2)

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            45.0, 44.7, 48.5, 50.2, 42.8, 50.0, 54.0, 52.0, 58.0, 60.0)

# Create the main data frame
datac <- data.frame(ID, var1, var2)

# Calculate summary statistics
mean_var1c <- round(mean(var1, na.rm=T),2)
mean_var2c <- round(mean(var2, na.rm=T),2)
sd_var1c <- round(sd(var1, na.rm=T),2)
sd_var2c <- round(sd(var2, na.rm=T),2)

cor_test <- cor.test(var1, var2, use = "complete.obs")
correlc <- round(cor_test$estimate, 2)
if(cor_test$p.value < .001) {
  p_valuec <- "< .001"} else {
  p_valuec <- paste0("= ", round(cor_test$p.value, 3)) }


p1 <- ggplot(data = datac, mapping = aes(x = var1, y = var2)) +
  geom_point(size = 7, color = "#00A9E0", alpha = 0.6) + # Increase point size
  geom_smooth(method = "lm", se = FALSE, color = "#8DC63F", size = 3.5, alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
  sm_statCorr(color = "#8DC63F", text_size = 8) + 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 


# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)


# Calculate summary statistics
mean_var1m <- round(mean(var1, na.rm=T),2)
mean_var2m <- round(mean(var2, na.rm=T),2)
sd_var1m <- round(sd(var1, na.rm=T),2)
sd_var2m <- round(sd(var2, na.rm=T),2)

var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            rep(mean_var2m, 10))

mean_var2m <- round(mean(var2, na.rm=T),2)
sd_var2m <- round(sd(var2, na.rm=T),2)

# Create the main data frame
datam <- data.frame(ID, var1, var2)

cor_test <- cor.test(var1, var2, use = "complete.obs")
correlm <- round(cor_test$estimate, 2)
if(cor_test$p.value < .001) {
  p_valuem <- "< .001"} else {
  p_valuem <- paste0("= ", round(cor_test$p.value, 3)) }

p2 <- ggplot(data = datam, mapping = aes(x = var1, y = var2)) +
  geom_point(size = 7, color = "#00A9E0", alpha = 0.6) + # Increase point size
  geom_smooth(method = "lm", se = FALSE, color = "#8DC63F", size = 3.5) +
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
  sm_statCorr(color = "#8DC63F", text_size = 8) + 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 

# Combine p1 and p2, adding p2 on top of p1

p3 <- ggplot(data = datam, mapping = aes(x = var1, y = var2)) +
  
  sm_statCorr(color = "white", text_size = 8) + 
  
  geom_point(data = datac, size = 7, color = "#00A9E0", alpha = 0.05) + # Increase point size
  #geom_smooth(data = datac, method = "lm", se = FALSE, color = "#8DC63F", size = 3.5, alpha = 0.2) +
  stat_smooth(data = datac, geom='line', method = "lm", alpha=0.08, se=FALSE, color = "#8DC63F", size = 3.5) +
  
  geom_point(data = datam, size = 7, color = "#00A9E0", alpha = 0.8) +
  #geom_smooth(data = datam, method = "lm", se = FALSE, color = "#8DC63F", size = 3.5) +
  stat_smooth(data = datam, geom='line', method = "lm", alpha=1, se=FALSE, color = "#8DC63F", size = 3.5) +
  
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 

p3

```
<small><small><small>.highlight[Complete Data Reference: *R* = `r correlc`, *p* `r p_valuec`]</small></small></small>

]

.pull-left-10[
```{r}

# for table
datam$var1 <- sprintf("%.2f", datam$var1)
datam$var2 <- ifelse(is.na(datam$var2), "", sprintf("%.2f", datam$var2))

# Create the table with kable and apply styling
knitr::kable(
  datam,
  format = "html",
  align = "ll",
  escape = FALSE
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    font_size = 13.5, full_width = TRUE
  )
```

]

---


#  Last Observation Carried Forward

  - Replace missing values with last observed score
  - Simple and again seems intuitive 
  - Can drastically inflate associations 

---

class: center, middle

.pull-right-60[

## LOCF `r emo::ji("see_no_evil")` `r emo::ji("hear_no_evil")` `r emo::ji("speak_no_evil")`

```{r, out.width = "450px", out.height = "425px", echo = F, message=FALSE, warning=FALSE, dpi = 300}
library(dplyr)
library(ggplot2)
library(smplot2)

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            45.0, 44.7, 48.5, 50.2, 42.8, 50.0, 54.0, 52.0, 58.0, 60.0)

# Create the main data frame
datac <- data.frame(ID, var1, var2)

# Calculate summary statistics
mean_var1c <- round(mean(var1, na.rm=T),2)
mean_var2c <- round(mean(var2, na.rm=T),2)
sd_var1c <- round(sd(var1, na.rm=T),2)
sd_var2c <- round(sd(var2, na.rm=T),2)

cor_test <- cor.test(var1, var2, use = "complete.obs")
correlc <- round(cor_test$estimate, 2)
if(cor_test$p.value < .001) {
  p_valuec <- "< .001"} else {
  p_valuec <- paste0("= ", round(cor_test$p.value, 3)) }


p1 <- ggplot(data = datac, mapping = aes(x = var1, y = var2)) +
  geom_point(size = 7, color = "#00A9E0", alpha = 0.6) + # Increase point size
  geom_smooth(method = "lm", se = FALSE, color = "#8DC63F", size = 3.5, alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
  sm_statCorr(color = "#8DC63F", text_size = 8) + 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 


# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)


# Calculate summary statistics
mean_var1m <- round(mean(var1, na.rm=T),2)
mean_var2m <- round(mean(var2, na.rm=T),2)
sd_var1m <- round(sd(var1, na.rm=T),2)
sd_var2m <- round(sd(var2, na.rm=T),2)

var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)

mean_var2m <- round(mean(var2, na.rm=T),2)
sd_var2m <- round(sd(var2, na.rm=T),2)

# Create the main data frame
datam <- data.frame(ID, var1, var2)

cor_test <- cor.test(var1, var2, use = "complete.obs")
correlm <- round(cor_test$estimate, 2)
if(cor_test$p.value < .001) {
  p_valuem <- "< .001"} else {
  p_valuem <- paste0("= ", round(cor_test$p.value, 3)) }

p2 <- ggplot(data = datam, mapping = aes(x = var1, y = var2)) +
  geom_point(size = 7, color = "#00A9E0", alpha = 0.6) + # Increase point size
  geom_smooth(method = "lm", se = FALSE, color = "#8DC63F", size = 3.5) +
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
  sm_statCorr(color = "#8DC63F", text_size = 8) + 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 

# Combine p1 and p2, adding p2 on top of p1

p3 <- ggplot(data = datam, mapping = aes(x = var1, y = var2)) +
  
  sm_statCorr(color = "white", text_size = 8) + 
  
  geom_point(data = datac, size = 7, color = "#00A9E0", alpha = 0.05) + # Increase point size
  #geom_smooth(data = datac, method = "lm", se = FALSE, color = "#8DC63F", size = 3.5, alpha = 0.2) +
  stat_smooth(data = datac, geom='line', method = "lm", alpha=0.08, se=FALSE, color = "#8DC63F", size = 3.5) +
  
  geom_point(data = datam, size = 7, color = "#00A9E0", alpha = 0.8) +
  #geom_smooth(data = datam, method = "lm", se = FALSE, color = "#8DC63F", size = 3.5) +
  stat_smooth(data = datam, geom='line', method = "lm", alpha=1, se=FALSE, color = "#8DC63F", size = 3.5) +
  
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 

p3

```
<small><small><small>.highlight[Complete Data Reference: *R* = `r correlc`, *p* `r p_valuec`]</small></small></small>

]

.pull-left-10[
```{r}

# for table
datam$var1 <- sprintf("%.2f", datam$var1)
datam$var2 <- ifelse(is.na(datam$var2), "", sprintf("%.2f", datam$var2))

# Create the table with kable and apply styling
knitr::kable(
  datam,
  format = "html",
  align = "ll",
  escape = FALSE
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    font_size = 13.5, full_width = TRUE
  )
```

]

---

#  Single Imputation

  - Replace missing values with the predicted score
  - Imputing from observed values 
  - Does not capture variability (regression line)
  - Inflates associations

---

class: center, middle

.pull-right-60[

## One Imputation `r emo::ji("thinking")`

```{r, out.width = "450px", out.height = "425px", echo = F, message=FALSE, warning=FALSE, dpi = 300}
library(dplyr)
library(ggplot2)
library(smplot2)

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            45.0, 44.7, 48.5, 50.2, 42.8, 50.0, 54.0, 52.0, 58.0, 60.0)

# Create the main data frame
datac <- data.frame(ID, var1, var2)

# Calculate summary statistics
mean_var1c <- round(mean(var1, na.rm=T),2)
mean_var2c <- round(mean(var2, na.rm=T),2)
sd_var1c <- round(sd(var1, na.rm=T),2)
sd_var2c <- round(sd(var2, na.rm=T),2)

cor_test <- cor.test(var1, var2, use = "complete.obs")
correlc <- round(cor_test$estimate, 2)
if(cor_test$p.value < .001) {
  p_valuec <- "< .001"} else {
  p_valuec <- paste0("= ", round(cor_test$p.value, 3)) }


p1 <- ggplot(data = datac, mapping = aes(x = var1, y = var2)) +
  geom_point(size = 7, color = "#00A9E0", alpha = 0.6) + # Increase point size
  geom_smooth(method = "lm", se = FALSE, color = "#8DC63F", size = 3.5, alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
  sm_statCorr(color = "#8DC63F", text_size = 8) + 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 


# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)


# Calculate summary statistics
mean_var1m <- round(mean(var1, na.rm=T),2)
mean_var2m <- round(mean(var2, na.rm=T),2)
sd_var1m <- round(sd(var1, na.rm=T),2)
sd_var2m <- round(sd(var2, na.rm=T),2)

# Fit a linear regression model using mar data
model <- lm(var2 ~ var1)

# Extract the intercept and slope
intercept <- coef(model)[1]
slope <- coef(model)[2]

# Calculate the predicted values for var2
var2[11:20] <- intercept + slope * var1[11:20]

mean_var2m <- round(mean(var2, na.rm=T),2)
sd_var2m <- round(sd(var2, na.rm=T),2)

# Create the main data frame
datam <- data.frame(ID, var1, var2)

cor_test <- cor.test(var1, var2, use = "complete.obs")
correlm <- round(cor_test$estimate, 2)
if(cor_test$p.value < .001) {
  p_valuem <- "< .001"} else {
  p_valuem <- paste0("= ", round(cor_test$p.value, 3)) }

p2 <- ggplot(data = datam, mapping = aes(x = var1, y = var2)) +
  geom_point(size = 7, color = "#00A9E0", alpha = 0.6) + # Increase point size
  geom_smooth(method = "lm", se = FALSE, color = "#8DC63F", size = 3.5) +
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
  sm_statCorr(color = "#8DC63F", text_size = 8) + 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 

# Combine p1 and p2, adding p2 on top of p1

p3 <- ggplot(data = datam, mapping = aes(x = var1, y = var2)) +
  
  sm_statCorr(color = "white", text_size = 8) + 
  
  geom_point(data = datac, size = 7, color = "#00A9E0", alpha = 0.05) + # Increase point size
  #geom_smooth(data = datac, method = "lm", se = FALSE, color = "#8DC63F", size = 3.5, alpha = 0.2) +
  stat_smooth(data = datac, geom='line', method = "lm", alpha=0.08, se=FALSE, color = "#8DC63F", size = 3.5) +
  
  geom_point(data = datam, size = 7, color = "#00A9E0", alpha = 0.8) +
  #geom_smooth(data = datam, method = "lm", se = FALSE, color = "#8DC63F", size = 3.5) +
  stat_smooth(data = datam, geom='line', method = "lm", alpha=1, se=FALSE, color = "#8DC63F", size = 3.5) +
  
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 

p3

```
<small><small><small>.highlight[Complete Data Reference: *R* = `r correlc`, *p* `r p_valuec`]</small></small></small>

]

.pull-left-10[
```{r}

# for table
datam$var1 <- sprintf("%.2f", datam$var1)
datam$var2 <- ifelse(is.na(datam$var2), "", sprintf("%.2f", datam$var2))

# Create the table with kable and apply styling
knitr::kable(
  datam,
  format = "html",
  align = "ll",
  escape = FALSE
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    font_size = 13.5, full_width = TRUE
  )
```

]

---

#  Stochastic Imputation

  - Replace missing values with regression-based estimate and add variability
  - Better but does not capture the variability across imputations 

---

class: center, middle

.pull-right-60[

## Stochastic `r emo::ji("nerd_face")`

```{r, out.width = "450px", out.height = "425px", echo = F, message=FALSE, warning=FALSE, dpi = 300}

library(dplyr)
library(ggplot2)
library(smplot2)

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            45.0, 44.7, 48.5, 50.2, 42.8, 50.0, 54.0, 52.0, 58.0, 60.0)

# Create the main data frame
datac <- data.frame(ID, var1, var2)

# Calculate summary statistics
mean_var1c <- round(mean(var1, na.rm=T),2)
mean_var2c <- round(mean(var2, na.rm=T),2)
sd_var1c <- round(sd(var1, na.rm=T),2)
sd_var2c <- round(sd(var2, na.rm=T),2)

cor_test <- cor.test(var1, var2, use = "complete.obs")
correlc <- round(cor_test$estimate, 2)
if(cor_test$p.value < .001) {
  p_valuec <- "< .001"} else {
  p_valuec <- paste0("= ", round(cor_test$p.value, 3)) }


p1 <- ggplot(data = datac, mapping = aes(x = var1, y = var2)) +
  geom_point(size = 7, color = "#00A9E0", alpha = 0.6) + # Increase point size
  geom_smooth(method = "lm", se = FALSE, color = "#8DC63F", size = 3.5, alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
  sm_statCorr(color = "#8DC63F", text_size = 8) + 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 


# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)


# Calculate summary statistics
mean_var1m <- round(mean(var1, na.rm=T),2)
mean_var2m <- round(mean(var2, na.rm=T),2)
sd_var1m <- round(sd(var1, na.rm=T),2)
sd_var2m <- round(sd(var2, na.rm=T),2)


set.seed(4681)  # Setting a seed for reproducibility

# Create a function to generate a frame for the scatter plot
generate_frame <- function(i) {
  
  # Create sample data
  random_error <- rnorm(20, mean = 0, sd = 3.51)

  # Calculate the predicted values for hrql2 with random error
  model <- lm(var2 ~ var1)
  
  # Extract the intercept and slope
  intercept <- coef(model)[1]
  slope <- coef(model)[2]
  
  # Calculate the predicted values for var2
  var2[11:20] <- intercept + slope * var1[11:20] + random_error[11:20]
  
  mean_var2m <- round(mean(var2, na.rm=T),2)
  sd_var2m <- round(sd(var2, na.rm=T),2)

  # Create the main data frame
  datam <- data.frame(ID, var1, var2)

  cor_test <- cor.test(var1, var2, use = "complete.obs")
  correlm <- round(cor_test$estimate, 2)
    if(cor_test$p.value < .001) {
       p_valuem <- "< .001"} else {
       p_valuem <- paste0("= ", round(cor_test$p.value, 3)) }
  
  
  # Create the scatter plot with regression line
p2 <- ggplot(data = datam, mapping = aes(x = var1, y = var2)) +
  sm_statCorr(color = "white", text_size = 8) + 
  
  geom_point(data = datac, size = 7, color = "#00A9E0", alpha = 0.05) + # Increase point size
  #geom_smooth(data = datac, method = "lm", se = FALSE, color = "#8DC63F", size = 3.5, alpha = 0.2) +
  stat_smooth(data = datac, geom='line', method = "lm", alpha=0.08, se=FALSE, color = "#8DC63F", size = 3.5) +
  
  geom_point(data = datam, size = 7, color = "#00A9E0", alpha = 0.8) +
  #geom_smooth(data = datam, method = "lm", se = FALSE, color = "#8DC63F", size = 3.5) +
  stat_smooth(data = datam, geom='line', method = "lm", alpha=1, se=FALSE, color = "#8DC63F", size = 3.5) +
  theme_classic() +
  scale_x_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) +  
  scale_y_continuous(breaks = seq(20, 70, by = 10), limits = c(20, 70)) + 
  sm_statCorr(color = "#8DC63F", text_size = 8) + 
   theme(
    plot.title = element_text(hjust = 0.5, size = 24),  # Increase title font size
    axis.title = element_text(size = 18),  # Increase axis label font size
    axis.text = element_text(size = 18),  # Increase tick label font size
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "#A7A9AC", size = 1.5),
    axis.ticks = element_line(size = 1.2),
    axis.ticks.length = unit(0.25, "cm"),
    text = element_text(color = "#4B4F54")
  ) 
  
  return(p2)
}

# Save the scatter plot as a GIF
animation_filename <- here::here("slides", "img", "scatter_plot.gif")

p3 <- animation::saveGIF({
  for (i in 1:5) {
    p <- generate_frame(i)
    print(p)
  }
}, movie.name = animation_filename, interval = 0.50, ani.width = 500, ani.height = 450)

```

`r knitr::include_graphics(animation_filename)`

<small><small><small>.highlight[Complete Data Reference: *R* = `r correlc`, *p* `r p_valuec`]</small></small></small>

]

.pull-left-10[
```{r out.width = "100%", out.height= "100%", echo = F, message=FALSE, warning=FALSE, dpi = 300}

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)

set.seed(4681)  # Setting a seed for reproducibility

library(gridExtra) 

# Create a function to generate a frame for the scatter plot
generate_frame <- function(i) {
  
  # Create sample data
  random_error <- rnorm(20, mean = 0, sd = 3.51)

  # Calculate the predicted values for hrql2 with random error
  model <- lm(var2 ~ var1)
  
  # Extract the intercept and slope
  intercept <- coef(model)[1]
  slope <- coef(model)[2]
  
  # Calculate the predicted values for var2
  var2[11:20] <- intercept + slope * var1[11:20] + random_error[11:20]
  
  mean_var2m <- round(mean(var2, na.rm=T),2)
  sd_var2m <- round(sd(var2, na.rm=T),2)

  # Create the main data frame
  datam <- data.frame(ID, var1, var2)

  # for table
  datam$var1 <- sprintf("%.2f", datam$var1)
  datam$var2 <- ifelse(is.na(datam$var2), "", sprintf("%.2f", datam$var2))

# Generate table as a grob
table_grob <- tableGrob(datam, rows = NULL, 
                        theme = ttheme_minimal(
                          core = list(
                            fg_params = list(hjust = 0, x = 0.1, fontsize = 16),
                            bg_params = list(fill = c("white", "#D9F3F1")),  # Alternating row colors
                            # Adding white borders around cells
                            lwd = 2,  # Line width
                            col = "white"  # Border color
                          ),
                          colhead = list(
                            fg_params = list(fontsize = 16, col = "#000000"),
                            bg_params = list(fill = "#D9F3F1")  # Optional: Set header background color
                          )
                        ))

  # Convert table grob to ggplot object
  p <- ggplot() + 
    annotation_custom(table_grob) +
    theme_void()  # Remove any ggplot default theme elements

  return(p)

}

# Save the scatter plot as a GIF
table_filename <- here::here("slides", "img", "scatter_table.gif")

p3 <- animation::saveGIF({
  for (i in 1:5) {
    p <- generate_frame(i)
    print(p)
  }
}, movie.name = table_filename, interval = 0.50, ani.width = 300, ani.height = 600)


```

`r knitr::include_graphics(table_filename)`

]

---

class: center, middle

# Any questions?



