---
title: "Handling Missing Data in Research: A Practical Guide"
subtitle  : "<small>full information maximum likelihood (FIML, direct-ml)</small>"
author: "`r paste(params$author, collapse = ' & ')`"
date: "`r paste0(params$location, ', ', params$date)`"
params:
  author: ["Author Name"]
  date: "Placeholder Date"
  location: "Placeholder Location"
---
exclude: true

```{r child = "_setup.Rmd"}
```

---

## ML estimation

- ML identifies the population parameters that are most likely given the observed data
- A likelihood (or log-likelihood) function is used to quantify how well the proposed parameters explain the observed data.
- ML requires a population distribution (normal)

---

class: center, middle

## ML estimation `r emo::ji("smirk")` 

A density function gives the shape of the normal curve

$$ 
L_i = {\frac{1}{\sqrt{2 \pi \sigma^2}} e^{-.5\frac{\color{#8DC63F}{(Y_i - \mu)}^2 }{\color{#8DC63F}{\sigma}^2} }} 
$$
.small[
$L_i$ (the likelihood) gives the relative probability that $Y_i$ came from a normal distribution with a particular mean and variance.
]

---

class: center, middle

.pull-right-60[

```{r, out.width = "450px", out.height = "425px", echo = F, message=FALSE, warning=FALSE, dpi = 300}
library(dplyr)
library(ggplot2)
library(smplot2)

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)

data1 <- data.frame(ID, var1)

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)

# Create the main data frame
datam <- data.frame(ID, var1, var2)

#--- fiml 

mu <- mean(data1$var1, na.rm=T)
sigma <- sd(data1$var1, na.rm=T)

data1$Li <- round(1 / sqrt(2 * pi * sigma^2) * exp(((data1$var1 - mu)^2 / sigma^2) * -0.5), 6) 

```

```{r, message = FALSE, warning = FALSE}

library(dplyr)
library(knitr)
library(kableExtra)

# Find the largest value in the last column
data1 <- data1 %>%
  mutate(across(
    .cols = ncol(data1),
    .fns = ~ ifelse(. == max(., na.rm = TRUE), 
                    paste0("<b style='font-size:16px;'>", ., "</b>"), 
                    .)
  ))

# Rename the last column to "L_i" with a subscript
colnames(data1)[ncol(data1)] <- "<div style='text-align: center;'>L<sub><i>i</i></sub></div>"

# Create the table with kable and apply styling
knitr::kable(
  data1,
  format = "html",
  align = "ll",
  escape = FALSE
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    font_size = 13.5, full_width = TRUE
  )

```

]

.pull-left-10[

### ML estimation

.small[
Applying the density function gives the relative probability $\text{(}L_i\text{)}$ of each score from this normal distribution. 


**complete data:** 

var1 $\text{( }\mu$ = `r sprintf("%.2f", mu)`, $\sigma$ = `r sprintf("%.2f", sigma)` $\text{)}$
]

]

---

class: center, middle

```{r, out.width = "600px", out.height = "600px", echo = F, message=FALSE, warning=FALSE, dpi = 300}

# Load necessary library
library(ggplot2)

data1$likelihood <- 1 / sqrt(2 * pi * sigma^2) * exp(-0.5 * ((data1$var1 - mu)^2) / sigma^2)

df <- data.frame(x = seq(25, 75, by = 0.5))
df$likelihood <- 1 / sqrt(2 * pi * sigma^2) * exp(-0.5 * ((df$x - mu)^2) / sigma^2)

# Define the x positions for vertical lines and find the closest mu values
select_x1 <- 45
select_x2 <- 62


# Create the plot
ggplot(df, aes(x = x, y = likelihood)) +
  geom_line(color = "#00A9E0", size = 3.50) +
  
  scale_y_continuous(breaks = seq(0, max(df$likelihood)), limits = c(0, max(df$likelihood)+.01), expand = c(0, 0))+

  # Left 
  annotate("text", x = select_x1 - 7, y = df$likelihood[df$x == select_x1], 
           label = bquote(italic(L)[.(sprintf("%.0f", select_x1))] == .(sprintf("%.2f", df$likelihood[df$x == select_x1]))), 
           color = "#4B4F54", size = 6)+
  geom_segment(aes(x = select_x1, y = min(likelihood), xend = select_x1, yend = df$likelihood[df$x == select_x1]), 
               color = "#A7A9AC", linetype = "dashed", size = 1) +
  geom_point(aes(x = select_x1, y = df$likelihood[df$x == select_x1]), color = "#8DC63F", size = 6) +
  
  # Right
  annotate("text", x = select_x2 + 7, y = df$likelihood[df$x == select_x2], 
           label = bquote(italic(L)[.(sprintf("%.0f", select_x2))] == .(sprintf("%.2f", df$likelihood[df$x == select_x2]))),
           color = "#4B4F54", size = 6) +
  geom_segment(aes(x = select_x2, y = min(likelihood), xend = select_x2, yend = df$likelihood[df$x == select_x2]), 
               color = "#A7A9AC", linetype = "dashed", size = 1) +
  geom_point(aes(x = select_x2, y = df$likelihood[df$x == select_x2]), color = "#8DC63F", size = 6) +

  labs(title = "", x = "Mean (var1)", y = "Likelihood") +
  theme(
    plot.margin = unit(c(1, 1, 1, 1), "cm"),
    panel.background = element_blank(),
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, margin = margin(b = 15)),
    axis.line = element_line(color = "black"),
    axis.title = element_text(size = 22, color = "black", face = "bold"),
    axis.text = element_text(size = 14, color = "black"),
    axis.text.x = element_text(margin = margin(t = 10)),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.title.x = element_text(size = 14, color = "black"),
    axis.ticks.y = element_blank()
  )

```

---

class: center, middle

.pull-right-40[

```{r, out.width = "600px", out.height = "600px", echo = F, message=FALSE, warning=FALSE, dpi = 300}
library(dplyr)
library(ggplot2)
library(smplot2)

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)

data1 <- data.frame(ID, var1)

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)

# Create the main data frame
datam <- data.frame(ID, var1, var2)

#--- fiml 

mu <- mean(data1$var1, na.rm=T)
sigma <- sd(data1$var1, na.rm=T)

data1$Li <- round(1 / sqrt(2 * pi * sigma^2) * exp(((data1$var1 - mu)^2 / sigma^2) * -0.5), 6)
data1$logLi <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data1$var1 - mu)^2 / sigma^2) * -0.5)), 6)

options(scipen = 999)
smallval <- prod(data1$Li)
smallval <- as.character(smallval)
options(scipen = 0)

sumfit <- round(sum(data1$logLi),2)

```

```{r, message = FALSE, warning = FALSE}

library(dplyr)
library(knitr)
library(kableExtra)

# Find the largest value in the last column
data1 <- data1 %>%
  mutate(across(
    .cols = ncol(data1),
    .fns = ~ ifelse(. == max(., na.rm = TRUE), 
                    paste0("<b style='font-size:16px;'>", ., "</b>"), 
                    .)
  ))

# Rename the last column to "L_i" with a subscript
colnames(data1)[ncol(data1)-1] <- "<div style='text-align: center;'>L<sub><i>i</i></sub></div>"
colnames(data1)[ncol(data1)] <- "<div style='text-align: center;'>logL<sub><i>i</i></sub></div>"


# Create the table with kable and apply styling
knitr::kable(
  data1,
  format = "html",
  align = "ll",
  escape = FALSE
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    font_size = 13.5, full_width = TRUE
  )

```

]

.pull-left-40[

### Maximum Likelihood

.small[
Multiple each $(L_i)$ to get sample likelihood.

<small><small><small>`r smallval`</small></small></small> 
<small><small>Fit of this data to $\mu$ = `r sprintf("%.2f", mu)`, $\sigma$ = `r sprintf("%.2f", sigma)`</small></small> 

<small>**To avoid small numbers, we take the log of the likelihood.**</small>  

Add each $logL_i$ to get sample loglikelihood.

`r sumfit`

]

]

---

class: center, middle

```{r, out.width = "600px", out.height = "600px", echo = F, message=FALSE, warning=FALSE, dpi = 300}
library(dplyr)
library(ggplot2)
library(smplot2)

# Data for each row in the table
ID <- 1:20
var1 <- c(36.6, 41.8, 42.6, 43.1, 43.4, 44.2, 44.9, 46.3, 48.6, 49.0, 
            50.0, 51.6, 54.6, 54.8, 55.7, 57.2, 57.6, 60.3, 60.9, 65.3)

data1 <- data.frame(ID, var1)

#--- fiml 

mu1 <- 30
mu2 <- 40
mu3 <- 50
mu4 <- 60
mu5 <- 70

sigma <- sd(data1$var1, na.rm=T)

data1$logLi1 <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data1$var1 - mu1)^2 / sigma^2) * -0.5)), 3)
data1$logLi2 <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data1$var1 - mu2)^2 / sigma^2) * -0.5)), 3)
data1$logLi3 <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data1$var1 - mu3)^2 / sigma^2) * -0.5)), 3)
data1$logLi4 <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data1$var1 - mu4)^2 / sigma^2) * -0.5)), 3)
data1$logLi5 <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data1$var1 - mu5)^2 / sigma^2) * -0.5)), 3)

#data1$logLi <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data1$var1 - mu)^2 / sigma^2) * -0.5)), 6)

sumfit1 <- round(sum(data1$logLi1), 2)
sumfit2 <- round(sum(data1$logLi2), 2)
sumfit3 <- round(sum(data1$logLi3), 2)
sumfit4 <- round(sum(data1$logLi4), 2)
sumfit5 <- round(sum(data1$logLi5), 2)

new_row <- c("", "", "", "", "", "", "")

data1 <- rbind(data1, new_row)

new_row <- c("", "", sumfit1, sumfit2, sumfit3, sumfit4, sumfit5)

data1 <- rbind(data1, new_row)

```

```{r, message = FALSE, warning = FALSE}

library(dplyr)
library(knitr)
library(kableExtra)

# Rename the last column to "L_i" with a subscript
colnames(data1)[ncol(data1)-4] <- paste0("\u03BC = ", mu1)
colnames(data1)[ncol(data1)-3] <- paste0("\u03BC = ", mu2)
colnames(data1)[ncol(data1)-2] <- paste0("\u03BC = ", mu3)
colnames(data1)[ncol(data1)-1] <- paste0("\u03BC = ", mu4)
colnames(data1)[ncol(data1)] <- paste0("\u03BC = ", mu5)

# Bold the entire last row
data1 <- data1 %>%
  mutate(across(
    everything(),  # Apply to all columns
    .fns = ~ ifelse(row_number() == n(), 
                    paste0("<b style='font-size:12px;'>", ., "</b>"), 
                    .)
  ))

# Create the table with kable and apply styling
knitr::kable(
  data1,
  format = "html",
  align = "ll",
  escape = FALSE
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    font_size = 11, full_width = TRUE
  )

```

---

class: center, middle

.pull-right-40[

```{r, out.width = "600px", out.height = "600px", echo = F, message=FALSE, warning=FALSE, dpi = 300}
library(dplyr)
library(ggplot2)
library(smplot2)

# Data for each row in the table
ID <- 1:20
var2 <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0, 
            NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)

data2 <- data.frame(ID, var2)

mu1 <- 30
mu2 <- 40
mu3 <- 50
mu4 <- 60
mu5 <- 70

sigma <- sd(data2$var2, na.rm=T)

data2$logLi1 <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data2$var2 - mu1)^2 / sigma^2) * -0.5)), 3)
data2$logLi2 <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data2$var2 - mu2)^2 / sigma^2) * -0.5)), 3)
data2$logLi3 <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data2$var2 - mu3)^2 / sigma^2) * -0.5)), 3)
data2$logLi4 <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data2$var2 - mu4)^2 / sigma^2) * -0.5)), 3)
data2$logLi5 <- round(log(1 / sqrt(2 * pi * sigma^2) * exp(((data2$var2 - mu5)^2 / sigma^2) * -0.5)), 3)

sumfit1 <- round(sum(data2$logLi1, na.rm=T), 2)
sumfit2 <- round(sum(data2$logLi2, na.rm=T), 2)
sumfit3 <- round(sum(data2$logLi3, na.rm=T), 2)
sumfit4 <- round(sum(data2$logLi4, na.rm=T), 2)
sumfit5 <- round(sum(data2$logLi5, na.rm=T), 2)

new_row <- c("", "", "", "", "", "", "")

data2 <- rbind(data2, new_row)

new_row <- c("", "", sumfit1, sumfit2, sumfit3, sumfit4, sumfit5)

data2 <- rbind(data2, new_row)

```

```{r, message = FALSE, warning = FALSE}

library(dplyr)
library(knitr)
library(kableExtra)

# Rename the last column to "L_i" with a subscript
colnames(data2)[ncol(data2)-4] <- paste0("\u03BC = ", mu1)
colnames(data2)[ncol(data2)-3] <- paste0("\u03BC = ", mu2)
colnames(data2)[ncol(data2)-2] <- paste0("\u03BC = ", mu3)
colnames(data2)[ncol(data2)-1] <- paste0("\u03BC = ", mu4)
colnames(data2)[ncol(data2)] <- paste0("\u03BC = ", mu5)

# Bold the entire last row
data2 <- data2 %>%
  mutate(across(
    everything(),  # Apply to all columns
    .fns = ~ ifelse(row_number() == n(), 
                    paste0("<b style='font-size:12px;'>", ., "</b>"), 
                    .)
  ))

data2 <- data2 %>%
  dplyr::mutate(across(2:7, ~ ifelse(is.na(.), "", 
                                     if (is.numeric(.)) sprintf("%.2f", .) else as.character(.))))

# Create the table with kable and apply styling
knitr::kable(
  data2,
  format = "html",
  align = "ll",
  escape = FALSE
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    font_size = 11, full_width = TRUE
  )

```
]

.pull-left-40[
<small><small><small>*Possible* population means for **var2**</small></small></small> 

```{r, out.width = "600px", out.height = "600px", echo = F, message=FALSE, warning=FALSE, dpi = 300}

library(ggplot2)
library(gganimate)
library(gifski)
library(grDevices)

# Define the new range for x-axis (means of variable var1)
means <- seq(30, 70, length.out = 50)

# Create a data frame for the animation
data <- data.frame(
  x = rep(seq(30, 70, length.out = 100), length(means)),
  mean = rep(means, each = 100)
)

# Compute Log-Likelihood (example formula)
# Replace dnorm density with log-likelihood (e.g., sum of log(density) or log probabilities)
# Assuming standard deviation = 5
data$y <- dnorm(data$x, mean = data$mean, sd = 5, log = TRUE)

custom_ticks <- c(40.0, 40.0, 35.0, 43.0, 42.6, 39.0, 45.0, 45.2, 50.0, 40.0)

# Create the plot with ggplot2
p <- ggplot(data, aes(x = x, y = y, group = mean)) +
  geom_line(color = "#00A9E0", linewidth = 2) +
  labs(
    title = "",
    x = "Mean (var2)",
    y = "Log-Likelihood"
  ) +
  # Add vertical lines for custom tick values
  geom_vline(
    xintercept = custom_ticks,
    color = "#8DC63F",
    linetype = "dashed",
    size = 0.8
  ) +
  theme(
    plot.margin = unit(c(1, 1, 1, 1), "cm"),
    panel.background = element_blank(),
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, margin = margin(b = 15)),
    axis.line = element_line(color = "black"),
    axis.title = element_text(size = 22, color = "black", face = "bold"),
    axis.text = element_text(size = 24, color = "black"),
    axis.text.x = element_text(margin = margin(t = 10)),
    axis.title.x = element_text(size = 24, color = "black")
  ) +
  transition_states(mean, transition_length = 1, state_length = 1) +
  ease_aes('linear')

animation_filename <- here::here("slides", "img", "moving_plot.gif")

# Animate and save as a .gif
#anim <- animate(p, nframes = 200, fps = 40, width = 900, height = 600, renderer = gifski_renderer())
#anim_save(animation_filename, animation = anim)

```

`r knitr::include_graphics(animation_filename)`

.small[
<small>Audition different parameters to quantify how well the proposed values explain the observed data.</small> 

<small>Green dotted lines represent observed values for var2.</small> 

<small>.highlight[*Note*. Listwise var2 *M* = 41.98.]</small>
]

]

---

class: left, middle

# Reporting `r emo::ji("page_facing_up")` `r emo::ji("page_facing_up")` `r emo::ji("page_facing_up")`

<small>**Step 3**. Report methods and software.</small>

.small[
> To address the issue of incomplete data, we used full information maximum likelihood (FIML) in R using the lavaan package (version 0.6-18). To account for non-normality in the data, we applied the robust standard error estimator by specifying the robust = TRUE option in the lavaan function. Additionally, we followed Graham’s (2003) approach for incorporating auxiliary variables into the analysis. Specifically, we included five auxiliary variables: age, sex, race, education, and income to handle potential bias introduced by missing data.   
]
<small><small>
**Extended Reporting**. Provide supplemental materials for details and code.
</small></small>

---

## Myths `r emo::ji("lying_face")`

- FIML "fills in" missing data
- Only for SEM models
- Must have MAR to use MI/FIML
- Guaranteed better than MI
- Exogenous variables are always included 
- No need with large samples

---

class: center, middle

# Any questions?



